Estimating the Initial Susceptible Fraction in Dengue Dynamics

### 2014-09-29 draft_lm1.tex
Flavio's corrections

The data used to fit the model consists of weekly incidence from 1996 to 2014 
in the city of Rio de Janeiro, Brazil.
From this series, we calculated the effective reproduction number, $R_t$, 
according to the expression given by Nishiura et al.~\cite{nishiura}.
The method assumes the disease counts $Y(t)$ are Poisson random variables such 
that $R_t$ is 
\[ R_t = \left(\frac{Y(t + 1)}{Y(t)}\right)^{1/n}\]
where $n$ is the ratio between the length of reporting interval and the mean 
generation time of the disease.
Here we obtain the estimates of $R_t$ using a moving window of three weeks, 
which roughly corresponds to the generation time of dengue, and thus $n = 1$.

@@

As the epidemic progresses, the effective transmission  rate changes and is
\begin{equation} 
 \label{eq:effbeta}
 \beta(t) = \frac{R_t\cdot\tau}{S_0}
\end{equation}

We parameterize our model with the varying transmission using the system of 
ordinary differential equations
\begin{align}
 \frac{dS}{dt} &= -\beta(t)SI \\
 \frac{dI}{dt} &= \beta(t)SI - \tau I&\\
 \frac{dR}{dt} &= \tau I&
\end{align}
where $S + I + R = 1 \: \forall\: t$. % \in [T_0, T_1]$. 
This is a rather simplified model, in which, for instance, the vector is left 
out.
Also, although there probably are multiple circulating serotypes, our approach
does not discriminate between them.

@@

For the fitting procedure the time series were divided into epidemic windows 
that corresponded to significant raises in incidence and normalized to lie on 
the $[0,1]$ interval.

Also, as the numerical integration routine needs $\beta(t)$ to be continuous, we 
used linear interpolation to obtain values of $R_t$ for any time point.

@@

To approximate the posterior in~(\ref{eq:S0post}) we use Markov chain Monte 
Carlo techiniques implemented in the Bayesian inference with Python 
(BIP)~\cite{pone2011} available at 
\url{http://code.google.com/p/bayesian-inference/}.
BIP uses a Differential Evolution Adaptive Metropolis (DREAM)~\cite{vrugt2008} 
scheme that efficiently samples from high-dimensional joint distributions using 
multiple adaptive chains running in parallel with delayed rejection.
The chains were run until $5000$ samples were obtained after discarding $500$ 
burn-in samples.
Convergence of the parallel chains was verified at every 100 iterations by the 
calculation of the Gelman-Rubins' R convergence diagnostic~\cite{brooks1998}.
% Results and Discussion can be combined.
\section*{Results}
% We only support three levels of headings, please do not create a heading level below \subsubsection.
%\subsection*{Subsection 1}
\section*{Discussion}
% Do NOT remove this, even if you are not including acknowledgments.
\section*{Acknowledgments}
LMC is grateful to Dr. Leonardo Bastos for useful discussions on the 
posterior inference for $R(t)$.

>>>>

The data used to fit the model consists of weekly incidence from 1996 to 2014 
in the city of Rio de Janeiro, Brazil.
From this series, we calculated the effective reproduction number, $R_t$, 
according to the expression given by Nishiura et al.~\cite{nishiura}.
The method assumes the disease counts $Y(t)$ are Poisson random variables such 
that $R_t$ is 
\begin{equation}
 \label{eq:Rt}
 R_t = \left(\frac{Y(t + 1)}{Y(t)}\right)^{1/n}
\end{equation}

where $n$ is the ratio between the length of reporting interval and the mean 
generation time of the disease.
Here we obtain the estimates of $R_t$ using a moving window of three weeks, 
which roughly corresponds to the generation time of dengue, and thus $n = 1$.

@@

As the epidemic progresses, the effective transmission  rate changes and is
\begin{equation} 
 \label{eq:effbeta}
 \beta(t) = \frac{R_t\cdot\tau}{S_0}
\end{equation}
where $R_t$ is the effective reproductive number as defined in~\ref{eq:Rt}.
We parameterize our model with the varying transmission using the system of 
ordinary differential equations
\begin{align}
 \frac{dS}{dt} &= -\beta(t)SI \\
 \frac{dI}{dt} &= \beta(t)SI - \tau I&\\
 \frac{dR}{dt} &= \tau I&
\end{align}
where $S + I + R = 1 \: \forall\: t$. % \in [T_0, T_1]$. 
This is a rather simplified model, in which, for instance, the vector is left 
out.
Also, although there probably are multiple circulating serotypes, our approach
does not discriminate between them due to the lack of serotype-specific data.

@@

For the fitting procedure the time series was divided into $J=10$ epidemic 
windows that corresponded to significant raises in incidence and normalized to 
lie on the $[0,1]$ interval.

Also, as the numerical integration routine needs to treat $\beta(t)$ as a
continuous function of time because of the variable step size, we used linear 
interpolation to obtain values of $R_t$ for any time point.

@@

To approximate the posterior in~(\ref{eq:S0post}) we use Markov chain Monte 
Carlo techiniques implemented in the Bayesian inference with Python 
(BIP)~\cite{pone2011} available at 
\url{http://code.google.com/p/bayesian-inference/}.
BIP uses a Differential Evolution Adaptive Metropolis (DREAM)~\cite{vrugt2008} 
scheme that efficiently samples from high-dimensional joint distributions using 
multiple adaptive chains running in parallel with delayed rejection.
In this study we used one chain per parameter, i.e, 3 chains for each run.
The chains were run until $5000$ samples were obtained after discarding $500$ 
burn-in samples.
Convergence of the parallel chains was verified at every 100 iterations by the 
calculation of the Gelman-Rubins' R convergence diagnostic~\cite{brooks1998}.
% Results and Discussion can be combined.
\section*{Results}
% We only support three levels of headings, please do not create a heading level below \subsubsection.
%\subsection*{Subsection 1}
\section*{Discussion}
% Do NOT remove this, even if you are not including acknowledgments.
\section*{Acknowledgments}
LMC is grateful to Dr. Leonardo Bastos for useful discussions on the 
posterior inference for $R(t)$.
